Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr

### 2017-07-24 cabios-template/draft.tex
draft2

\title[R packages for analyzing genome-wide data]{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author[Privé \textit{et~al}.]{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}
$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France and \\
$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

$^\ast$To whom correspondence should be addressed.

\date{~ }

\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge for the genetic community, severely slowing down genetic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} https://privefl.github.io/bigstatsr/ and https://privefl.github.io/bigsnpr/\\
\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\maketitle

\section{Introduction}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}. Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}. Then, implementation of GWAS analyses also depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple variants \cite[]{Vilhjalmsson2015,Euesden2015}. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a brake on data exploration.
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis, already implemented by the scientific community. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We are aware the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Usually, memory-mapping provides seamless and faster access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this binary PLINK format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which is 8 times less disk storage space than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}). The first function is a wrapper to PLINK and Beagle which takes PLINK files as input and return PLINK files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}. XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD, (Figure~\ref{fig:svd}).

@@

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance (e.g. association with a phenotype) and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions by the algorithm. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
Statistical tests based on linear and logistic regressions are also available. For the linear regression, for each SNP $j$, a T-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$ and $\hat{p} = \mathbb{P}(Y = 1)$.
\\
The hypothesis that is tested is $\beta^{(j)} = 0$ against the alternative $\beta^{(j)} \neq 0$.

@@

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Tibshirani1996,Zou2005,Friedman2010}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

@@

In this paper, two real datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same structures of Population and Linkage Disequilibrium as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset. \label{sec:rep}

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compared the performance of the packages against standard software. Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $1^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with our new method (see section~\ref{sec:impute}). We note that if we used a standard R matrix to store the genotypes, this data would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB.

We used bigstatsr and bigsnpr R functions to get the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly and takes only 20 lines of code.

The Celiac dataset is relatively small as compared to modern genetic cohorts. To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, on the whole dataset, we made a linear association test for each SNP, using the 10 first PCs as covariables.

@@

We first compared the GWAS and PRS computations with the R packages against PLINK 1.9 and EIGENSOFT 6.1.4.

For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g. pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}). However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.

On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 and the similar function implemented in bigstatsr, big\_randomSVD. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset (Section~\ref{sec:rep}). Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure~\ref{fig:bench-pca}). We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition (e.g. SmartPCA). FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).

@@

For the Celiac dataset, we compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}). When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure~\ref{fig:scores}). Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}) which is confirmed by Gini coefficients (measure of dispersion) of each squared loadings reported in Figure~\ref{fig:gini}, which are all around the theoretical value of $2 / pi$ (for gaussian loadings) and significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed.

@@

For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset. The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}). We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values. Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle (v21Jan17). Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours. We also show that our method's estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).

@@

>>>>

\title{Efficient management and analysis of large-scale genome-wide data with two R packages: bigstatsr and bigsnpr}
\author{Florian Privé\,$^{\text{ 1,}*}$, Hugues Aschard\,$^{\text{2,3}}$ and Michael G.B. Blum\,$^{\text{1,}*}$}

\date{~ }
\maketitle

\noindent$^{\text{\sf 1}}$Université Grenoble Alpes, CNRS, Laboratoire TIMC-IMAG, UMR 5525, France, \\
\noindent$^{\text{\sf 2}}$Centre de Bioinformatique, Biostatistique et Biologie Intégrative (C3BI), Institut Pasteur, Paris, France \\
\noindent$^{\text{\sf 3}}$Department of Epidemiology, Harvard T.H. Chan School of Public Health, Boston, Massachusetts, USA.

\noindent$^\ast$To whom correspondence should be addressed.

\newpage
\abstract{\textbf{Motivation:} Genome-wide datasets produced for association studies have dramatically increased in size over the past few years, with modern datasets commonly including millions of variants measured in dozens of thousands of individuals. This increase in data size is a major challenge severely slowing down genomic analyses. Specialized software for every part of the analysis pipeline have been developed to handle large genomic data. However, combining all these software into a single data analysis pipeline might be technically difficult. \\
\textbf{Results:} Here we present two R packages, bigstatsr and bigsnpr, allowing for management and analysis of large scale genomic data to be performed within a single comprehensive framework. To address large data size, the packages use memory-mapping for accessing data matrices stored on disk instead of in RAM. To perform data pre-processing and data analysis, the packages integrate most of the tools that are commonly used, either through transparent system calls to existing software, or through updated or improved implementation of existing methods. In particular, the packages implement a fast derivation of Principal Component Analysis, functions to remove SNPs in Linkage Disequilibrium, and algorithms to learn Polygenic Risk Scores on millions of SNPs. We illustrate applications of the two R packages by analysing a case-control genomic dataset for the celiac disease, performing an association study and computing Polygenic Risk Scores. Finally, we demonstrate the scalability of the R packages by analyzing a simulated genome-wide dataset including 500,000 individuals and 1 million markers on a single desktop computer. \\
\textbf{Availability:} \href{https://privefl.github.io/bigstatsr/}{https://privefl.github.io/bigstatsr/} and \href{https://privefl.github.io/bigsnpr/}{https://privefl.github.io/bigsnpr/}\\
\textbf{Contact:} \href{michael.blum@univ-grenoble-alpes.fr}{michael.blum@univ-grenoble-alpes.fr}\\
\textbf{Supplementary information:} Supplementary data are available at \textit{Bioinformatics}
online.}

\newpage
\section{Introduction}

@@

Consider first the basic tools necessary to perform a standard genome-wide analysis. Conversions between standard file formats has become a field by itself with several tools such as VCFtools, BCFtools and PLINK, available either independently or incorporated within large framework \cite[]{Danecek2011,Li2011,Purcell2007}. Similarly, quality control software for genome-wide analysis have been developed such as PLINK and Bioconductor GWASTools \cite[]{Gogarten2012}. Computation of principal components (PCs) of genotypes is commonly performed to account for population stratification in association studies and there are several software available, including EIGENSOFT (SmartPCA and FastPCA) and FlashPCA \cite[]{Price2006,Galinsky2016,Abraham2014a,Abraham2016a}. Then, implementation of GWAS analyses also depends on the genotype format, e.g. ProbABEL or SNPTEST for dosage data \cite[]{Aulchenko2010,Marchini2010}.  Finally, there exists a range of tools for Polygenic Risk Scores (PRSs) such as LDpred and PRSice, which provide prediction for quantitative traits or disease risks based on multiple genetic variants \cite[]{Vilhjalmsson2015,Euesden2015}. As a result, one has to make extensive bash/perl/R/python scripts to link these software together and convert between multiple file formats, involving many file manipulations and conversions. Overall, this might be a brake on data exploration.
%Overall, this means that researchers are usually restricted on how they can manipulate and analyse the data they have access to.

Secondly, increasing size of genetic datasets is the source of major computational challenges and many analytical tools would be restricted by the amount of memory (RAM) available on computers. This is particularly a burden for commonly used analysis languages such as R, Python and Perl. Solving the memory issues for these languages would give access to a broad range of tools for data analysis that have been already implemented. Hopefully, strategies have been developed to avoid loading large datasets in RAM. For storing and accessing matrices, memory-mapping is very attractive because it is seamless and usually much faster to use than direct read or write operations. Storing large matrices on disk and accessing them via memory-mapping is available in R through ``big.matrix'' objects implemented in the R package bigmemory \cite[]{Kane2013}. Thanks to this matrix-like format, algorithms in R/C++ can be developed or adapted for large genotype data.

@@

We are aware of the software library SNPFile that uses memory-mapped files to store and efficiently access genotype data, coded in C++ \cite[]{Nielsen2008}. With the two packages we developed, we made this solution available in R and in C++ via package Rcpp \cite[]{Eddelbuettel2011}. The major advantage of manipulating genotype data within R, almost as it were a standard matrix in memory, is the possibility of using most of the other tools that have been developed in R \cite[]{R2017}. For example, we provide sparse multivariate linear models and an efficient algorithm for Principal Component Analysis (PCA) based on adaptations from R packages biglasso, sparseSVM and RSpectra \cite[]{RSpectra2016,Zeng2017}.

Usually, memory-mapping provides seamless and faster access than standard read/write operations. When some element is needed, a small chunk of the genotype matrix, containing this element, is accessed in memory. When the system needs more memory, some chunks of the matrix are freed from the memory in order to make space for others. All this is managed by the Operating System so that it is seamless and efficient. It means that if you use the same chunks of data repeatedly, it will be very fast the second time you access it, the third time and so on. Of course, if the memory size of your computer is larger than the size of the dataset, the file could fit entirely in memory and every second access would be fast.

@@

Fast read/write operations from/to bed/bim/fam PLINK files are available so that one should convert data to this format. In bigsnpr, we provide R functions that use system calls to PLINK for the conversion and the Quality Control steps (Figure~\ref{fig:qc}). PLINK files are then read into a ``bigSNP'' object, which contains the genotype ``big.matrix'', a data frame with information on samples and another data frame with information on SNPs. We also provide another function which could be used to read from tabular-like text files in order to create a genotype in the format ``big.matrix''.

We developed a special ``big.matrix'' object, called ``BM.code'', that can be used to seamlessly store up to 256 arbitrary different values, while having a relatively efficient storage. Indeed, each element is stored on one byte which requires 8 times less disk storage than double-precision numbers but 4 times more space than the binary PLINK format ``.bed''. With these 256 values, the matrix can store genotype calls and missing values (4 values), best guess genotypes (3 values) and genotype dosages (likelihoods) rounded to two decimal places (201 values).

Because it is an important part of the preprocessing, we provide two functions for imputing missing values of genotyped SNPs (Figure~\ref{fig:impute}). The first function is a wrapper to PLINK and Beagle which takes bed files as input and return bed files without missing values, and should therefore be used before reading the data in R \cite[]{Browning2009}. The second function is a new algorithm we developed in order to have a fast imputation method without losing much of imputation accuracy. This algorithm doesn't use phasing and is very fast, as previous Machine Learning Approaches for genetic imputation \cite[]{Wang2012}. It only relies on some local XGBoost models. XGBoost is an optimized distributed gradient boosting library that can be used in R and provides some of the best results in machine learning competitions \cite[]{Chen2016}. XGBoost builds decision trees that can detect nonlinear interactions, partially reconstructing phase so that it seems well suited for imputing genotype matrices. For each SNP, we provide an estimation of imputation error by separating non-missing data into training/test sets. The training set is used to build a model for predicting missing data. The prediction model is then evaluated on the test set for which we know the true genotype values, which gives an unbiased estimator of the number of individuals that have been wrongly imputed for that particular SNP.

@@

For computing Principal Components (PCs) of a large-scale genotype matrix, we provide several functions related to SNP thinning and two functions, for computing a partial Singular Value Decomposition (SVD), one based on eigenvalue decomposition, big\_SVD, and the other on randomized projections, big\_randomSVD (Figure~\ref{fig:svd}).

@@

A variant of pruning is clumping. Clumping is useful if a statistic is available to sort the SNPs by importance, e.g. association with a phenotype, and for discarding SNPs in LD with a more associated SNP relatively to the phenotype of interest. Furthermore, we advise to always use clumping instead of pruning (by using the minor allele frequency as the statistic of importance, which is the default) because, in some particular cases, pruning can leave regions of the genome without any representative SNP at all\footnote{https://goo.gl/T5SJqM}.

The third option that is generally combined with pruning or clumping consists of removing SNPs in long-range LD regions \cite[]{Price2008}. Long-range LD regions for the human genome are available as an online table that our packages can use to discard SNPs in long-range LD regions while computing PCs\footnote{https://goo.gl/8TngVE}. However, such table is human specific and could also be population specific, so we developed an algorithm that automatically detects these regions and removes them. This algorithm consists in the following steps: first, PCA is performed using a subset of SNP remaining after clumping, then outliers SNPs are detected using Mahalanobis distance as implemented in the R package pcadapt \cite[]{Luu2017}. Finally, the algorithm considers that consecutive outlier SNPs are in long-range LD regions. Indeed, a long-range LD region would cause SNPs in this region to have strong consecutive weights (loadings) in the PCA. This algorithm is implemented in function snp\_autoSVD and will be referred by this name in the rest of the paper.

@@

Any test statistic that is based on counts could be easily implemented because we provide fast counting summaries. Among these tests, the Armitage trend test and the MAX3 test statistic are already provided for binary outcome \cite[]{Zheng2012}. 
Statistical tests based on linear and logistic regressions are also available. For the linear regression, for each SNP $j$, a t-test is performed on $\beta^{(j)}$ where $\hat{y} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $K$ is the number of Principal Components and $L$ is the number of other covariates (such as the age and gender). Similarly, for the logistic regression, for each SNP $j$, a Z-test is performed on $\beta^{(j)}$ where $\log{\frac{\hat{p}}{1-\hat{p}}} = \alpha^{(j)} + \beta^{(j)} SNP^{(j)} + \gamma_1^{(j)} PC_1 + \cdots + \gamma_K^{(j)} PC_K + \delta_1^{(j)} COV_1 + \cdots + \delta_K^{(j)} COV_L$, and $\hat{p} = \mathbb{P}(Y = 1)$ and $Y$ denotes the binary phenotype.

@@

Secondly, the two R packages also implement multivariate models to compute risk scores that do not use univariate summary statistics but instead train a model on all the SNPs and covariables at once, optimally accounting for correlation between predictors \cite[]{Abraham2012}. The currently available models are linear and logistic regressions and Support Vector Machine (SVM). These models include lasso and elastic-net regularizations, which reduce the number of  predictors (SNPs) included in the predictive models \cite[]{Tibshirani1996,Zou2005,Friedman2010}. Package bigstatsr provides a fast implementation of these models by using efficient rules to discard most of the predictors \cite[]{Tibshirani2012}. The implementation of these algorithms is based on modified versions of functions available in the R packages sparseSVM and biglasso \cite[]{Zeng2017}. These modifications allow to include covariates in the models and to use these algorithms on the special type of ``big.matrix'' called ``BM.code'' used in bigsnpr (see section~\ref{sec:preprocess}).

@@

In this paper, two datasets are analyzed: the celiac disease cohort and the POPRES datasets \cite[]{Dubois2010,Nelson2008}. The Celiac dataset is composed of 15,283 individuals of European ancestry genotyped on 295,453 SNPs. The POPRES dataset is composed of 1385 individuals of European ancestry genotyped on 447,245 SNPs.

For comparing computation times of different software, we replicated all the individuals in the Celiac dataset 5 and 10 times in order to have larger datasets while keeping the same population structure and pattern of Linkage Disequilibrium as the original dataset. To assess scalibility of our algorithms for a biobank-scale genotype dataset, we formed another dataset of 500,000 individuals and 1 million SNPs, also by replicating the Celiac dataset. \label{sec:rep}

@@

We present the results for three different analyses. First, we illustrate the application of R packages bigstatsr and bigsnpr. Secondly, we compare the performance of the R packages to the performance obtained with PLINK and FastPCA (EIGENSOFT). Thirdly, we present results of the two new methods implemented in these packages, one method for the automatic detection and removal of long-range LD regions in PCA and another for the imputation of missing genotypes. We use three types of data: a case-control cohort for the celiac disease, the European population cohort POPRES and simulated datasets using real genotypes from the Celiac cohort. We compare the performance on two computers, a desktop computer with 64GB of RAM and 12 cores, and a laptop with only 8GB of RAM and 4 cores. For the functions that enable parallelism, we use half of the cores available on the corresponding computer.

@@

We performed an association study and computed a polygenic risk score for the Celiac cohort. 
The data was preprocessed following steps from figure~\ref{fig:qc}, removing individuals and SNPs which had more than 5\% of missing values, non-autosomal SNPs, SNPs with a minor allele frequency lower than 0.05 or a p-value for the Hardy-Weinberg exact test lower than $10^{-10}$, and finally, removing one individual in each pair of individuals with a proportion of alleles shared IBD greater than 0.08 \cite[]{Purcell2007}. 
For the POPRES dataset, this resulted in 1382 individuals and 344,614 SNPs with no missing value.
For the Celiac dataset, this resulted in 15,155 individuals and 281,122 SNPs with an overall genotyping rate of 99.96\% that was then imputed with the XGBoost method. If we used a standard R matrix to store the genotypes, this data would take 32GB of memory. On the disk, the ``.bed'' file takes 1GB and the ``.bk'' file (storing the ``big.matrix'') takes 4GB only.

We used bigstatsr and bigsnpr R functions to compute the first Principal Components (PCs) of a genotype matrix and to visualize them (Figure~\ref{fig:pca}). We then performed a Genome-Wide Association Study (GWAS) investigating how Single Nucleotide Polymorphisms (SNPs) are associated with the celiac disease, while accounting for population structure with PCs, and plotted the results as a Manhattan plot (Figure~\ref{fig:gwas}). As illustrated in the supplementary data, the whole pipeline is user-friendly and requires only 20 lines of R code.

%The Celiac dataset is relatively small as compared to modern genetic cohorts. 
To illustrate the scalability of the two R packages, we performed a GWAS analysis on 500K individuals and 1M SNPs. The GWAS analysis completed in less than 5 hours using the aforementioned desktop computer. The GWAS analysis was composed of three main steps. First, we removed SNPs in long-range LD regions and used SNP clumping, leaving 93,083 SNPs. Then, the 10 first PCs were computed on the 500K individuals and these remaining SNPs. Finally, on the whole dataset, we made a linear association test for each SNP, using the 10 first PCs as covariables.

@@

We first compared the GWAS and PRS computations obtained with the R packages to the ones obtained with PLINK 1.9 and EIGENSOFT 6.1.4.
For most functions, multithreading is not available yet in PLINK, nevertheless, PLINK-specific algorithms that use bitwise parallelism (e.g. pruning) are still faster than the parallel algorithms reimplemented in package bigsnpr (Table~\ref{tab:bench-gwas}). Overall, the computations with our two R packages for an association study and a polygenic risk score are of the same order of magnitude as when using PLINK and EIGENSOFT (Tables~\ref{tab:bench-gwas} and~\ref{tab:bench-prs}). However, the whole analysis pipeline makes use of R calls only; there is no need to write temporary files and functions have parameters which enable subsetting of the genotype matrix without having to copy it.

On our desktop computer, we compared the computation times of FastPCA, FlashPCA2 to the similar function big\_randomSVD implemented in bigstatsr. For each comparison, we used the 93,083 SNPs which were remaining after pruning and we computed 10 PCs. We used the datasets of growing size simulated from the Celiac dataset. Overall, our function big\_randomSVD showed to be twice as fast as FastPCA and FlashPCA2 and almost 10 times as fast when using parallelism with 6 cores (Figure~\ref{fig:bench-pca}). We also compared results in terms of precision by comparing squared correlation between approximated PCs and ``true'' PCs provided by an exact singular value decomposition obtained with SmartPCA. FastPCA, FlashPCA2 and bigstatsr infer the true first 6 PCs but the squared correlation between true PCs and approximated ones decreases for larger PCs when using FastPCA whereas it remains larger than 0.999 when using FlashPCA2 or bigstatsr (Figure~\ref{fig:prec-pca}).

@@

For the Celiac dataset, we compared results of PCA obtained when using snp\_autoSVD and when computing PCA without removing any long range LD region (only clumping at $R^2 > 0.2$). When not removing any long range LD region, we show that PC4 and PC5 corresponds to a long-range LD region in chromosome 8 (Figures~\ref{fig:scores} and~\ref{fig:loadings}). When automatically removing some long-range LD regions with snp\_autoSVD, we show that PC4 and PC5 are now only reflecting population structure (Figure~\ref{fig:scores}). Moreover, loadings are more equally distributed among SNPs (Figure~\ref{fig:loadings}). This is confirmed by Gini coefficients (measure of dispersion) of each squared loadings that are significantly smaller when computing SVD with snp\_autoSVD than when no long-range LD region is removed (Figure~\ref{fig:gini}).

@@

For the fast imputation method based on XGBoost, we compared  the imputation accuracy and computation times on the POPRES dataset. The minor allele frequencies (MAFs) are approximately uniformly distributed between 0.05 and 0.5 (Figure~\ref{fig:maf}). We introduced missing values using a Beta-binomial distribution (Figure~\ref{fig:NA}) resulting in approximately 3\% of missing values. Imputation was compared between function snp\_fastImpute of package bigsnpr and Beagle 4.1 (version of January 21, 2017). Overall, our method made 4.7\% of imputation errors whereas Beagle made only 3.1\% but it took Beagle 14.6 hours to complete while our method only took 42 minutes (20 times less). For the Celiac dataset, our method took less than 6 hours to complete whereas Beagle didn't finish imputing chromosome 1 in 48 hours. We also show that the estimation of the number of imputation errors is accurate (Figure~\ref{fig:error-impute}).

@@

\renewcommand{\thefigure}{S\arabic{figure}}
\setcounter{figure}{0}
\renewcommand{\thetable}{S\arabic{table}}
\setcounter{table}{0}
